---
title: "Error-in-Variables Regression"
author: "Aaron R. Caldwell"
date: "Last Updated: `r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Error-in-Variables Regression}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Background

Error-in-variables (EIV) models are useful tools to account for measurement error in the independent variable. For studies of agreement, this is particularly useful where there are paired measurements of the paired measurements (X & Y) of the same underlying value (e.g., two assays of the same analyte).

Deming regression is one of the simplest forms of of EIV models promoted by [W. Edwards Deming](https://en.wikipedia.org/wiki/W._Edwards_Deming)^[Deming was a titan of the fields of statistics and engineering and I would highly recommend reading some of his academic work and books]. The first to detail the method were @adcock1878 followed by @kummell1879 and @koopmans1936. The name comes from the popularity of Deming's book [@deming1943], and within the field of clinical chemistry, the procedure was simply referred to as "Deming regression" (e.g., @linnet1990). 

# Joint Confidence Regions

A key, but currently experimental, enhancement to the `dem_reg` function is the addition of **joint confidence regions** for the slope and intercept parameters. Traditional confidence intervals treat each parameter separately, but in regression analysis, these parameters are correlated. 

Joint confidence regions account for this correlation by creating an elliptical region in the (intercept, slope) parameter space. This approach, promoted by @sadler2010, provides several advantages:

1. **Higher Statistical Power**: The ellipse typically requires 20-50% fewer samples than traditional confidence intervals to detect the same bias
2. **Accounts for Parameter Correlation**: When the measurement range is narrow, slope and intercept are highly negatively correlated
3. **More Appropriate Test**: Testing against a point (e.g., slope=1, intercept=0) is naturally done with a region, not separate intervals

The power advantage is most pronounced when the ratio of maximum to minimum X values is small (< 10:1), which is common in clinical method comparisons.

# Basic Deming Regression

## Simple Deming Regression

We can start by creating some fake data to work with.

```{r}
library(SimplyAgree)

dat = data.frame(
  x = c(7, 8.3, 10.5, 9, 5.1, 8.2, 10.2, 10.3, 7.1, 5.9),
  y = c(7.9, 8.2, 9.6, 9, 6.5, 7.3, 10.2, 10.6, 6.3, 5.2)
)
```

Also, we will assume, based on historical data, that the measurement error ratio is equal to 2.

The data can be run through the `dem_reg` function and the results printed. 

```{r}
dem1 = dem_reg(y ~ x,
               data = dat,
               error.ratio = 2,
               weighted = FALSE)
dem1
```

The resulting regression line can then be plotted. The subtitle now indicates whether the identity line is enclosed by the joint confidence region.

```{r}
plot(dem1, interval = "confidence")
```

## Visualizing the Joint Confidence Region

A new plotting method `plot_joint()` allows visualization of the confidence region in parameter space:

```{r}
plot_joint(dem1, 
           ideal_slope = 1, 
           ideal_intercept = 0,
           show_intervals = TRUE)
```

This plot shows:
- The **red ellipse**: Joint confidence region
- The **blue rectangle**: Traditional confidence intervals
- The **black dot**: Estimated (intercept, slope)
- The **green/red X**: Ideal point (whether enclosed or not)

Notice how the ellipse is smaller than the rectangle, especially in the directions that matter for detecting bias.

## Model Diagnostics

The assumptions of the Deming regression model, primarily normality and homogeneity of variance, can then be checked with the `check` method for Deming regression results. Both plots appear to be fine with regards to the assumptions.

```{r}
check(dem1)
```

# Weighted Deming Regression

For this example, I will rely upon the "ferritin" data from the `deming` R package.

```{r}
library(deming)
data('ferritin')

head(ferritin)
```

Let me demonstrate the problem with using simple Deming regression when the weights are helpful. When we look at the two plots below, we can see there is severe problem with using the "un-weighted" model.

```{r}
dem2 = dem_reg(
  old.lot ~ new.lot,
  data = ferritin,
  weighted = FALSE
)
summary(dem2)

plot(dem2)

plot_joint(dem2)

check(dem2)
```

Now, let us see what happens when `weighted` is set to TRUE.

```{r}
dem3 = dem_reg(
  x = "new.lot",
  y = "old.lot",
  data = ferritin,
  weighted = TRUE
)
summary(dem3)

plot(dem3)

plot_joint(dem3)

check(dem3)
```

The weighted model provides a better fit.


# Theoretical Details

## Calculative Approach

Deming regression assumes paired measures ($x_i, \space  y_i$) are each measured with error.

$$
x_i = X_i + \epsilon_i
$$

$$
y_i = Y_i + \delta_i
$$

We can then measure the relationship between the two variables with the following model.

$$
\hat Y_i = \beta_0 + \beta_1 \cdot \hat X_i
$$

Traditionally there are 2 null hypotheses

First, the intercept is equal to zero

$$
H_0: \beta_0 = 0 \space vs. \space H_1: \beta_0 \ne 0
$$

Second, that the slope is equal to one.

$$
H_0: \beta_1 = 1 \space vs. \space H_1: \beta_0 \ne 1
$$

## Joint Confidence Region

The joint $(1-\alpha) \times 100\%$ confidence region for $(\beta_0, \beta_1)$ is defined as:

$$
(\hat{\beta} - \beta_0)^T V^{-1} (\hat{\beta} - \beta_0) \leq \chi^2_{2,\alpha}
$$

Where:

- $\hat{\beta}$ = estimated parameters $(\hat{\beta}_0, \hat{\beta}_1)$
- $\beta_0$ = hypothesized parameters (e.g., $(0, 1)$ for identity)
- $V$ = variance-covariance matrix
- $\chi^2_{2,\alpha}$ = chi-square critical value with 2 degrees of freedom

This forms an ellipse in parameter space that accounts for the correlation between slope and intercept.

## Measurement Error

A Deming regression model also assumes the measurement error ($\sigma^2$) ratio is constant.

$$
\lambda = \frac{\sigma^2_\epsilon}{\sigma^2_\delta} 
$$

In `SimplyAgree`, the error ratio can be set with the `error.ratio` argument. It defaults to 1, but can be changed by the user. If replicate measures are taken, then the user can use the `id` argument to indicate which measures belong to which subject/participant. The measurement error, and the error ratio, will then be estimated from the data itself.

If the data was not measured in replicate then the error ratio ($\lambda$) can be estimated from the coefficient of variation (if that data is available) and the mean of x and y ($\bar x, \space \bar y$).

$$
\lambda = \frac{(CV_y \cdot \bar y)^2}{(CV_x \cdot \bar x)^2}
$$

## Weights

In some cases the variance of X and Y may increase proportional to the true value of the measure. In these cases, it may be prudent to use "weighted" Deming regression models. The weights used in `SimplyAgree` are the same as those suggested by @linnet1993.

$$
\hat w_i = \frac{1}{ [ \frac{x_i + \lambda \cdot y_i}{1 + \lambda}]^2}
$$

Weights can also be provided through the `weights` argument. If weighted Deming regression is not selected (`weighted = FALSE`), the weights for each observation is equal to 1.

The estimated mean of X and Y are then estimated as the following.

$$
\bar x_w =  \frac{\Sigma^{N}_{i=1} \hat w_i \cdot x_i}{\Sigma^{N}_{i=1} \hat w_i}
$$

$$
\bar y_w =  \frac{\Sigma^{N}_{i=1} \hat w_i \cdot y_i}{\Sigma^{N}_{i=1} \hat w_i}
$$

## Estimating the Slope and Intercept

First, there are 3 components ($v_x, \space v_y, \space cov_{xy}$)

$$
v_x = \Sigma_{i=1}^N \space \hat w_i \cdot (x_i- \bar x_w)^2
$$
$$
v_y = \Sigma_{i=1}^N \space \hat w_i \cdot (y_i- \bar y_w)^2
$$
$$
cov_{xy} = \Sigma_{i=1}^N \space \hat w_i \cdot  (x_i- \bar x_w) \cdot (y_i- \bar y_w)
$$

The slope ($b_1$) can then be estimated with the following equation.

$$
b_1 = \frac{(\lambda \cdot v_y - v_x) + \sqrt{(v_x-\lambda \cdot v_y)^2 + 4 \cdot \lambda \cdot cov_{xy}^2}}{2 \cdot \lambda \cdot cov_{xy}}
$$

The intercept ($b_0$) can then be estimated with the following equation.

$$
b_0 = \bar y_w - b_1 \cdot \bar x_w
$$

The standard errors of b1 and b0 are both estimated using a jackknife method (detailed by @linnet1990).

# References

